Taking uncertainty in word embedding bias estimation seriously: a bayesian approach

### 2021-07-21 paperDraft/paperDraft.Rmd
Adds small changes

---
title: "Taking uncertainty in word embedding bias estimation seriously: a  bayesian approach"
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: true
    df_print: kable 
    keep_tex: true
    includes:
      in_header:
        - Rafal_latex6.sty
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: [../references/bibliographyRafal.bib]
csl: [../references/apa-6th-edition.csl] 
---

@@

\textbf{1. Selection of attributes}

Such approaches, however, are not unproblematic. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
 We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.

@@

\textbf{3. Hiding the impact of uncertainty}

@@

\textbf{4. No word class distinction and no control group}

In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. One should investigate the religions separately as some of them may have stronger harmful associations that others. One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words. Later in the text we will explain in details reasons for introducing control groups. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

\textbf{5. Interpreting the results}

Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. @Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion). For now let us focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

@@

\textbf{6. The curse of dimensionality}

In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. According to the article 
[Curse of dimensionality at analyticsindiamag.com](https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/) as the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms. One may notice that more data should contribute to greater amount of information, but more information also means greater risk of noise and distractions in data. At the same time, modern solutions are often adapted to smaller dimensions and their results in higher ones are not intuitive, or may be prone to error.

Using cosine similarity in high dimensions in word embeddings may also be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data. Yet, according to some articles cosine similarity in high dimensions is not reliable enough as it may be the case that choosing words at random may result in similar values as when picking them consciously.

![curse of dimensionality,  number of dimensions on the x axis,  standard deviation of similarity on the y axis](../images/curseOfDimensionality.png)

>>>>

---
title: 'Taking uncertainty in word embedding bias estimation seriously: a  bayesian
  approach'
author: "Alicja Dobrzeniecka and Rafal Urbaniak"
output:
  pdf_document:
    number_sections: yes
    df_print: kable
    keep_tex: yes
    includes:
      in_header: Rafal_latex6.sty
  html_document:
    df_print: paged
classoption: dvipsnames,enabledeprecatedfontcommands
fontsize: 10pt
documentclass: scrartcl
urlcolor: blue
bibliography: ../references/bibliographyRafal.bib
csl: ../references/apa-6th-edition.csl
---

@@

Such approaches, however, are not unproblematic. One may cosinder the selection of attributes for the bias research. For one thing, the word lists are often fairly small, and the papers do not discuss the impact of word list sizes on the uncertainty involved. 
We will argue that the word list sizes are too small to yield strong conclusions, given the variances involved.

@@

@@

We also observed that in the paper there is no word class distinction and no control group. In the original paper, words from all three religions were compared against all of the stereotypes, which means that there was no distinction between cases in which the stereotype is associated with a given religion, as opposed to the situation in which it is associated with another one. Not all of the stereotypical words have to be considered as harmful for all of the religions. One should investigate the religions separately as some of them may have stronger harmful associations that others. One should also include control groups to have a way of comparing the stereotypical results with neutral or human-like words. Later in the text we will explain in details reasons for introducing control groups. In our analysis, we distinguish between stereotypes associated with a given group, stereotypes associated with different groups, and control groups: neutral words and stereotypes-free human predicates.

The interpretation of the results is also a challenge. Assuming for a moment that the value of multi-class cosine distance is correct, one may question the interpretation. @Manzini2019blackToCriminal summarize the averages of cosine distance per group (gender, race, religion). For now let us focus now on analyzing the values relating to religious biases. Here is the relevant fragment of table:

@@

One may consider the issue of the curse of dimensionality. In our case, the curse of dimensionality may take place when there is an increase in the volume of data that results in adding extra dimensions to the Euclidean space. As the number of features increases, it may be harder and harder to obtain useful information from the data using the available algorithms. Using cosine similarity in high dimensions in word embeddings may be prone to the curse of dimensionality. According to @Venkat2018Curse there are reasons to consider this phenomenon when searching for word similarities in higher dimensions. An experiment is conducted that aims at showing how the similarity values and variation change as the number of dimensions increases. The hypothesis made in the paper states that two things will happen as the number of dimensions increase. First, the effort required to measure cosine similarity will be greater, and two, the similarity between data will blur out and have less variation. The authors generate random points with increasing number of dimensions where each dimension of a data point is given a value between 0 and 1. Then they pick one vector at random from each dimension class and calculate the cosine similarity between the chosen vector and the rest of the data. Then they check how the variation of values changes as the number of dimensions increases. It seems like the more dimensions there are, the smaller the variance and therefore it is less obvious how to interpret the resulting cosine similarities. Maybe the scale should be adjusted to the number of dimensions and variance so that it still gives us sensible information about data.

### 2021-07-21 paperDraft/paperDraft.tex
Adds small changes

\title{A bayesian method of cosine-based word2vec bias}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

@@

One of the first measures in the discussion has been developed by
Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016). There, the gender
bias of a word \(w\) is understood as its projection on the gender
direction \(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\)
(the gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct gender bias is defined
as the average distance of the words in \(N\) from \(g\) (\(c\) is a
parameter determining how strict we want to be): \begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by Gonen \&
Goldberg (2019), who point out that while gender-direction might be an
indicator of bias, it is only one possible manifestation of it, and
reducing a projection of words might be insufficient. For instance,
``math'' and ``delicate'' might be in equal distance to both explicitly
gendered words while being closer to quite different stereotypical
attribute words. Further, the authors point out that most word pairs
preserve similarity under debiasing meant to minimize projection-based
bias.\footnote{Bolukbasi, Chang, Zou, Saligrama, \& Kalai (2016) use
  also another method which involves analogies and their evaluations by
  human users on Mechanical Turk. It is discussed and criticized in
  (Nissim, Noord, \& Goot, 2020).}

To measure bias in word embeddings, Caliskan, Bryson, \& Narayanan
(2017) proposed the Word Embedding Association Test (WEAT). The idea is
that the measure of biases between two sets of target words, \(X\) and
\(Y\), (we call them protected words) should be quantified in terms of
the cosine similarity between the protected words and attribute words
coming from two sets of stereotype attribute words, \(A\) and \(B\)
(we'll call them attributes). For instance, \(X\) might be a set of male
names, \(Y\) a set of female names, \(A\) might contain stereotypically
male-related career words, and \(B\) stereotypically female-related
family words. WEAT is a modification of the Implicit Association Test
(IAT) (Nosek, Banaji, \& Greenwald, 2002) used in psychology and uses
almost the same word sets, allowing for a \emph{prima facie} sensible
comparison with bias in humans. If the person's attitude towards given
pair of concept is to be interpreted as neutral, there should be no
noticeable task completion time difference, and the final value from the
formula should be around 0. Let \(f\) be a similarity measure (usually,
cosine similarity). The association difference for a term \(t\) is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align} \noindent then, the association difference between \(A\) a
\(B\) is: \begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
signifcance test, and the \(p\)-value obtained by bootstrapping: it is
the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
normalizing the difference in means as follows: \begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

Caliskan, Bryson, \& Narayanan (2017) show that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. Lauscher \& Glavas (2019) extended the
methodology to a multilingual and cross-lingual setting, arguing that
using Euclidean distance instead of similarity does not make much
difference, while the bias effects vary greatly across embedding models
(interestingly, with social media-text trained embeddings being less
biased than those based on Wikipedia).

A similar methodology is employed by Garg, Schiebinger, Jurafsky, \& Zou
(2018), who employ word embeddings trained on corpora from different
decades to study the shifts in various biases. For instance, to compute
the occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women
(e.g.~``she,'' ``female''), then calculate the Euclidean distance
between this mean vector and words for occupations. Then they take the
mean of these distances and subtract from it the analogously obtained
mean for the average vector of vector embeddings of words that represent
men. Formally they take the relative norm distance between \(X\) and
\(Y\) to be: \begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.

@@

\textbf{1. Selection of attributes}

The attributes are taken from different sources, there is no principled
justification for their choice. From our analysis it will become clear
that the list is rather uneven.

There is no mention of methodology for deciding on the number of
attributes necessary to decide a hypothesis on the given size of
dataset. There are however some ways to estimate sample sizes needed to
ensure that the results are meaningful if the effect is present. Our
research will show that the numbers used are rather insufficient.

\textbf{2. No attention to distributions and details}

The authors use the mean of mean average cosine similarities to measure
multi-class similarity between protected word and harmful stereotypes.
They average the results until they obtain one final value to represent
the mean cosine distance between protected word from a given class and
the attributes of that class. If one takes closer look at the individual
values that are taken for the calculations it turns out that there is a
bunch of outliers and surprisingly dissimilar words. We approach this
issue by providing the tables and visualizations of individual cosine
distances to make sure that we obtain a proper insight into the data.

\textbf{3. Hiding the impact of uncertainty}

@@

\textbf{4. No word class distinction and no control group}

In the original paper, words from all three religions were compared
against all of the stereotypes, which means that there was no
distinction between cases in which the stereotype is associated with a
given religion, as opposed to the situation in which it is associated
with another one. Not all of the stereotypical words have to be
considered as harmful for all of the religions. One should investigate
the religions separately as some of them may have stronger harmful
associations that others. One should also include control groups to have
a way of comparing the stereotypical results with neutral or human-like
words. Later in the text we will explain in details reasons for
introducing control groups. In our analysis, we distinguish between
stereotypes associated with a given group, stereotypes associated with
different groups, and control groups: neutral words and stereotypes-free
human predicates.

\textbf{5. Interpreting the results}

Assuming for a moment that the value of multi-class cosine distance is
correct, one may question the interpretation. Manzini, Lim, Tsvetkov, \&
Black (2019) summarize the averages of cosine distance per group
(gender, race, religion). For now let us focus now on analyzing the
values relating to religious biases. Here is the relevant fragment of
table:

@@

\textbf{6. The curse of dimensionality}

In our case, the curse of dimensionality may take place when there is an
increase in the volume of data that results in adding extra dimensions
to the Euclidean space. According to the article
\href{https://analyticsindiamag.com/curse-of-dimensionality-and-what-beginners-should-do-to-overcome-it/}{Curse
of dimensionality at analyticsindiamag.com} as the number of features
increases, it may be harder and harder to obtain useful information from
the data using the available algorithms. One may notice that more data
should contribute to greater amount of information, but more information
also means greater risk of noise and distractions in data. At the same
time, modern solutions are often adapted to smaller dimensions and their
results in higher ones are not intuitive, or may be prone to error.

Using cosine similarity in high dimensions in word embeddings may also
be prone to the curse of dimensionality. According to Venkat (2018)
there are reasons to consider this phenomenon when searching for word
similarities in higher dimensions. An experiment is conducted that aims
at showing how the similarity values and variation change as the number
of dimensions increases. The hypothesis made in the paper states that
two things will happen as the number of dimensions increase. First, the
effort required to measure cosine similarity will be greater, and two,
the similarity between data will blur out and have less variation. The
authors generate random points with increasing number of dimensions
where each dimension of a data point is given a value between 0 and 1.
Then they pick one vector at random from each dimension class and
calculate the cosine similarity between the chosen vector and the rest
of the data. Then they check how the variation of values changes as the
number of dimensions increases. It seems like the more dimensions there
are, the smaller the variance and therefore it is less obvious how to
interpret the resulting cosine similarities. Maybe the scale should be
adjusted to the number of dimensions and variance so that it still gives
us sensible information about data. Yet, according to some articles
cosine similarity in high dimensions is not reliable enough as it may be
the case that choosing words at random may result in similar values as
when picking them consciously.

\begin{figure}
\centering
\includegraphics{../images/curseOfDimensionality.png}
\caption{curse of dimensionality, number of dimensions on the x axis,
standard deviation of similarity on the y axis}
\end{figure}

@@

To illustrate let's employ the formulas used by Caliskan, Bryson, \&
Narayanan (2017) in a simple example. Conceptually, all such tests come
up with rather short lists of protected words and rather short lists of
stereotypical attributes. Clearly, these are not complete list. So let's
treat them as samples from richer pools of stereotypical predicates and
let's take the uncertainty and variation involved seriously.

@@

\noindent When you look at the datapoints, do you have the impression of
a strong bias here? We wouldn't think so. But now let's run the
calculations from (Caliskan, Bryson, \& Narayanan, 2017).

@@

\noindent This should make us pause. We know these were points randomly
drawn from distributions where there is no difference in means. Yet, the
calculated effect size is 1.82, whereas the largest effect size reported
by Caliskan, Bryson, \& Narayanan (2017) is 1.81!

@@

We noticed that in Manzini, Lim, Tsvetkov, \& Black (2019) words from
all three sub-classes (in the case of religion by sub-class we mean
Christianity, Judaism, Islam) were compared against all of the
stereotypes, which means that there was no distinction between cases in
which the stereotype is associated with a given sub-class, as opposed to
the situation in which it is associated with another one. Not all of the
stereotypical words have to be considered as harmful for all of the
sub-groups. One should investigate the sub-groups separately as some of
them may have stronger harmful associations than others.

@@

Our priors distributions for the Bayesian model reflect our knowledge
about the data. The distribution for mean distance (cosine distance) is
set to be between 0 and 2 with the mean equal to 1. We assume that most
of the cosine distances may be considered neutral and therefore their
similarity will be small. The distribution for prior for standard
deviation is mostly between 0 and 1 with mean around 0. Finally the
distribution for prior for coefficients is mostly between -1 and 1, with
mean around 0. One should remember that the cosine distance values have
range between 0 and 2 therefore the choice of the prior distributions
should reflect this constraint. As we want to verify as objectively as
possible the assumptions present in Manzini, Lim, Tsvetkov, \& Black
(2019) we decided to use fairly weak and slightly skeptical regularizing
priors. The models parameters will have a posterior distribution
obtained using either Hamiltionian Monte Carlo methods (STAN) available
through the rethinking package.

@@

Let's analyse dataset-level coeﬀicients together with their highest
posterior density intervals, for the three topic groups. We compare the
results for two different word emebeddings corpuses - RedditL2 and
GoogleNews. For Google embeddings the HPDIs for all class coeﬀicients
(associated, different, human, and none) include zero. This can lead one
to a conclusion that the impact for associated, different, human and
neutral attribute is, when averaged, quite similar. This indicates how
including the uncertainty may change the use and interpretation of
Manzini, Lim, Tsvetkov, \& Black (2019) MAC metric. It seems that if one
focuses only on differences between means of means, this is too
simplistic.

In the case of Reddit word embeddings the situation is similar although
HPDI is below 0 for Gender class when looking at \texttt{associated} and
\texttt{different} mean coeﬀicient. This can suggest that there is
indeed slightly stronger impact of these attributes. One should also
notice how in general associated and different coeﬀicients have quite
similar HPDI range, the highest observed absolute difference is equal to
only 0.1. This suggests that again the impact of associated attributes
and different ones is not clear at first sight. Finally, let's compare
the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race
dataset there is a minor shift (in absolute values the highest change is
equal to approximately 0.1) of the mean coeﬀicients towards zero.
However for Gender dataset there is no significant change. This is of
course a general look at the group coeﬀicients. Let's now analyze the
individual words.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-bolukbasi2016man}{}%
Bolukbasi, T., Chang, K.-W., Zou, J., Saligrama, V., \& Kalai, A.
(2016). Man is to computer programmer as woman is to homemaker?
Debiasing word embeddings. Retrieved from
\url{http://arxiv.org/abs/1607.06520}

\leavevmode\hypertarget{ref-Caliskan2017semanticsBiases}{}%
Caliskan, A., Bryson, J. J., \& Narayanan, A. (2017). Semantics derived
automatically from language corpora contain human-like biases.
\emph{Science}, \emph{356}(6334), 183--186. American Association for the
Advancement of Science ({AAAS}). Retrieved from
\url{https://doi.org/10.1126/science.aal4230}

\leavevmode\hypertarget{ref-Garg2018years}{}%
Garg, N., Schiebinger, L., Jurafsky, D., \& Zou, J. (2018). Word
embeddings quantify 100 years of gender and ethnic stereotypes.
\emph{Proceedings of the National Academy of Sciences}, \emph{115}(16),
E3635--E3644. Proceedings of the National Academy of Sciences. Retrieved
from \url{https://doi.org/10.1073/pnas.1720347115}

\leavevmode\hypertarget{ref-Gonen2019lipstick}{}%
Gonen, H., \& Goldberg, Y. (2019). Lipstick on a pig: {D}ebiasing
methods cover up systematic gender biases in word embeddings but do not
remove them. \emph{Proceedings of the 2019 conference of the north
{A}merican chapter of the association for computational linguistics:
Human language technologies, volume 1 (long and short papers)} (pp.
609--614). Minneapolis, Minnesota: Association for Computational
Linguistics. Retrieved from
\url{https://www.aclweb.org/anthology/N19-1061}

\leavevmode\hypertarget{ref-Lauscher2019multidimensional}{}%
Lauscher, A., \& Glavas, G. (2019). Are we consistently biased?
Multidimensional analysis of biases in distributional word vectors.
\emph{CoRR}, \emph{abs/1904.11783}. Retrieved from
\url{http://arxiv.org/abs/1904.11783}

\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019). Black is
to criminal as caucasian is to police: Detecting and removing multiclass
bias in word embeddings. Retrieved from
\url{http://arxiv.org/abs/1904.04047}

\leavevmode\hypertarget{ref-Nissim2020fair}{}%
Nissim, M., Noord, R. van, \& Goot, R. van der. (2020). Fair is better
than sensational: Man is to doctor as woman is to doctor.
\emph{Computational Linguistics}, \emph{46}(2), 487--497. {MIT} Press -
Journals. Retrieved from \url{https://doi.org/10.1162/coli_a_00379}

\leavevmode\hypertarget{ref-Nosek2002harvesting}{}%
Nosek, B. A., Banaji, M. R., \& Greenwald, A. G. (2002). Harvesting
implicit group attitudes and beliefs from a demonstration web site.
\emph{Group Dynamics: Theory, Research, and Practice}, \emph{6}(1),
101--115. American Psychological Association ({APA}). Retrieved from
\url{https://doi.org/10.1037/1089-2699.6.1.101}

\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
Venkat, N. (2018). The curse of dimensionality: Inside out.

>>>>

\title{Taking uncertainty in word embedding bias estimation seriously: a
bayesian approach}
\author{Alicja Dobrzeniecka and Rafal Urbaniak}
\date{}

@@

One of the first measures in the discussion has been developed by
(\textbf{bolukbasi2016man?}). There, the gender bias of a word \(w\) is
understood as its projection on the gender direction
\(\vec{w} \cdot (\overrightarrow{he} - \overrightarrow{she})\) (the
gender direction is the top principal compontent of ten gender pair
difference vectors). The underlying idea is that no bias is present if
non-explicitly gendered words are in equal distance to both elements in
all explicitly gender pairs. Given the (ideally) gender neutral words
\(N\) and the gender direction \(g\) the direct gender bias is defined
as the average distance of the words in \(N\) from \(g\) (\(c\) is a
parameter determining how strict we want to be): \begin{align}
\mathsf{directBias_c(N,g)} & = \frac{\sum_{w\in N}\vert \mathsf{cos}(\vec{w},g)\vert^c}{\vert N \vert }
\end{align}

The use of projections has been ciriticized for instance by
(\textbf{Gonen2019lipstick?}), who point out that while gender-direction
might be an indicator of bias, it is only one possible manifestation of
it, and reducing a projection of words might be insufficient. For
instance, ``math'' and ``delicate'' might be in equal distance to both
explicitly gendered words while being closer to quite different
stereotypical attribute words. Further, the authors point out that most
word pairs preserve similarity under debiasing meant to minimize
projection-based bias.\footnote{(\textbf{bolukbasi2016man?}) use also
  another method which involves analogies and their evaluations by human
  users on Mechanical Turk. It is discussed and criticized in
  (\textbf{Nissim2020fair?}).}

To measure bias in word embeddings,
(\textbf{Caliskan2017semanticsBiases?}) proposed the Word Embedding
Association Test (WEAT). The idea is that the measure of biases between
two sets of target words, \(X\) and \(Y\), (we call them protected
words) should be quantified in terms of the cosine similarity between
the protected words and attribute words coming from two sets of
stereotype attribute words, \(A\) and \(B\) (we'll call them
attributes). For instance, \(X\) might be a set of male names, \(Y\) a
set of female names, \(A\) might contain stereotypically male-related
career words, and \(B\) stereotypically female-related family words.
WEAT is a modification of the Implicit Association Test (IAT)
(\textbf{Nosek2002harvesting?}) used in psychology and uses almost the
same word sets, allowing for a \emph{prima facie} sensible comparison
with bias in humans. If the person's attitude towards given pair of
concept is to be interpreted as neutral, there should be no noticeable
task completion time difference, and the final value from the formula
should be around 0. Let \(f\) be a similarity measure (usually, cosine
similarity). The association difference for a term \(t\) is:
\begin{align}
s(t,A,B) & = \frac{\sum_{a\in A}f(t,a)}{\vert A\vert} - \frac{\sum_{b\in B}f(t,b)}{\vert B\vert}
\end{align} \noindent then, the association difference between \(A\) a
\(B\) is: \begin{align}
s(X,Y,A,B) & = \sum_{x\in X} s(x,A,B) -  \sum_{y\in Y} s(y,A,B)
\end{align} \noindent \(s(X,Y,A,B)\) is the statistic used in the
signifcance test, and the \(p\)-value obtained by bootstrapping: it is
the frequency of \(s(X_i,Y_i,A,B)>s(X,Y,A,B)\) for all equally sized
partitions \(X_i, Y_i\) of \(X\cup Y\). The effect size is computed by
normalizing the difference in means as follows: \begin{align}
bias(A,B) & = \frac{
\mu(\{s(x,A,B)\}_{x\in X}) -\mu(\{s(y,A,B)\}_{y\in Y}) 
}{
\sigma(\{s(w,A,B)\}_{w\in X\cup Y})
}
\end{align}

(\textbf{Caliskan2017semanticsBiases?}) show that significant
biases---thus measured--- similar to the ones discovered by IAT can be
discovered in word embeddings. (\textbf{Lauscher2019multidimensional?})
extended the methodology to a multilingual and cross-lingual setting,
arguing that using Euclidean distance instead of similarity does not
make much difference, while the bias effects vary greatly across
embedding models (interestingly, with social media-text trained
embeddings being less biased than those based on Wikipedia).

A similar methodology is employed by (\textbf{Garg2018years?}), who
employ word embeddings trained on corpora from different decades to
study the shifts in various biases. For instance, to compute the
occupational embeddings bias for women the authors first compute the
average vector of vector emeddings of words that represent women
(e.g.~``she,'' ``female''), then calculate the Euclidean distance
between this mean vector and words for occupations. Then they take the
mean of these distances and subtract from it the analogously obtained
mean for the average vector of vector embeddings of words that represent
men. Formally they take the relative norm distance between \(X\) and
\(Y\) to be: \begin{align}
\textsf{relative norm distance} & = \sum_{v_m\in M} \vert \vert v_m - v_X\vert \vert_2 - \vert v_m - v_Y\vert \vert_2
\end{align} \noindent where the norm used is Euclidean, and \(v_X\) and
\(x_Y\) are average vectors for sets \(X\) and \(Y\) respectively.

@@

Such approaches, however, are not unproblematic. One may cosinder the
selection of attributes for the bias research. For one thing, the word
lists are often fairly small, and the papers do not discuss the impact
of word list sizes on the uncertainty involved. We will argue that the
word list sizes are too small to yield strong conclusions, given the
variances involved.

Crucially, these approaches use means of mean average cosine
similarities to measure similarity between protected word and harmful
stereotypes. If one takes a closer look at the individual values that
are taken for the calculations, it turns out that there are quite a few
outliers and surprisingly dissimilar words. This issue will become
transparent when we inspect the visualizations of individual cosine
distances, following the idea that one of the first step to understand
data is to look at it.

@@

We also observed that in the paper there is no word class distinction
and no control group. In the original paper, words from all three
religions were compared against all of the stereotypes, which means that
there was no distinction between cases in which the stereotype is
associated with a given religion, as opposed to the situation in which
it is associated with another one. Not all of the stereotypical words
have to be considered as harmful for all of the religions. One should
investigate the religions separately as some of them may have stronger
harmful associations that others. One should also include control groups
to have a way of comparing the stereotypical results with neutral or
human-like words. Later in the text we will explain in details reasons
for introducing control groups. In our analysis, we distinguish between
stereotypes associated with a given group, stereotypes associated with
different groups, and control groups: neutral words and stereotypes-free
human predicates.

The interpretation of the results is also a challenge. Assuming for a
moment that the value of multi-class cosine distance is correct, one may
question the interpretation. Manzini, Lim, Tsvetkov, \& Black (2019)
summarize the averages of cosine distance per group (gender, race,
religion). For now let us focus now on analyzing the values relating to
religious biases. Here is the relevant fragment of table:

@@

One may consider the issue of the curse of dimensionality. In our case,
the curse of dimensionality may take place when there is an increase in
the volume of data that results in adding extra dimensions to the
Euclidean space. As the number of features increases, it may be harder
and harder to obtain useful information from the data using the
available algorithms. Using cosine similarity in high dimensions in word
embeddings may be prone to the curse of dimensionality. According to
Venkat (2018) there are reasons to consider this phenomenon when
searching for word similarities in higher dimensions. An experiment is
conducted that aims at showing how the similarity values and variation
change as the number of dimensions increases. The hypothesis made in the
paper states that two things will happen as the number of dimensions
increase. First, the effort required to measure cosine similarity will
be greater, and two, the similarity between data will blur out and have
less variation. The authors generate random points with increasing
number of dimensions where each dimension of a data point is given a
value between 0 and 1. Then they pick one vector at random from each
dimension class and calculate the cosine similarity between the chosen
vector and the rest of the data. Then they check how the variation of
values changes as the number of dimensions increases. It seems like the
more dimensions there are, the smaller the variance and therefore it is
less obvious how to interpret the resulting cosine similarities. Maybe
the scale should be adjusted to the number of dimensions and variance so
that it still gives us sensible information about data.

@@

To illustrate let's employ the formulas used by
(\textbf{Caliskan2017semanticsBiases?}) in a simple example.
Conceptually, all such tests come up with rather short lists of
protected words and rather short lists of stereotypical attributes.
Clearly, these are not complete list. So let's treat them as samples
from richer pools of stereotypical predicates and let's take the
uncertainty and variation involved seriously.

@@

\noindent When you look at the datapoints, do you have the impression of
a strong bias here? We wouldn't think so. But now let's run the
calculations from (\textbf{Caliskan2017semanticsBiases?}).

@@

\noindent This should make us pause. We know these were points randomly
drawn from distributions where there is no difference in means. Yet, the
calculated effect size is 1.82, whereas the largest effect size reported
by (\textbf{Caliskan2017semanticsBiases?}) is 1.81!

@@

One thing to notice is that in Manzini, Lim, Tsvetkov, \& Black (2019)
words from all three subclasses (in the case of religion by a subclass
we mean Christianity, Judaism, and Islam) were compared against all of
the stereotypes, which means that there was no distinction between cases
in which the stereotype is associated with a given sub-class, as opposed
to the situation in which it is associated with another one. Not all of
the stereotypical words have to be considered as harmful for all of the
sub-groups. One should investigate the sub-groups separately as some of
them may have stronger harmful associations than others.

@@

Our prior distributions for the Bayesian model reflect our first stab at
what we know or expect before we see the data. All our distributions are
fairly weak. The distribution for mean distance (cosine distance) is set
to fit (almost completely) between 0 and 2 with the mean equal to 1, in
line with the idea that most of the cosine distances may be considered
neutral and therefore their similarity will be around 0. That is, we
take it to be DISTRIBUTION. The prior distribution for standard
deviation is half-cauchy between 0 and 1, but unlikely to be as large as
1. Finally the prior distributions for the coefficients have most of
their mass between -1 and 1, with mean around 0. As we want to put
ourselves in skeptical shoes looking at Manzini, Lim, Tsvetkov, \& Black
(2019) and asking whether the data would indeed convince us of the
existence of bias, we decided to use slightly skeptical (mean 0), but
fairly weak (sd = ) as our regularizing priors. The model parameters
will have a posterior distribution obtained using either Hamiltionian
Monte Carlo methods (STAN) available through the rethinking package.

@@

Let's analyse dataset-level coefficients together with their highest
posterior density intervals, for the three topic groups. We compare the
results for two different word emebeddings corpuses - RedditL2 and
GoogleNews. For Google embeddings the HPDIs for all class coefficients
(associated, different, human, and none) include zero. This can lead one
to a conclusion that the impact for associated, different, human and
neutral attribute is, when averaged, quite similar. This indicates how
including the uncertainty may change the use and interpretation of
Manzini, Lim, Tsvetkov, \& Black (2019) MAC metric. It seems that if one
focuses only on differences between means of means, this is too
simplistic.

In the case of Reddit word embeddings the situation is similar although
HPDI is below 0 for Gender class when looking at \texttt{associated} and
\texttt{different} mean coefficient. This can suggest that there is
indeed slightly stronger impact of these attributes. One should also
notice how in general associated and different coefficients have quite
similar HPDI range, the highest observed absolute difference is equal to
only 0.1. This suggests that again the impact of associated attributes
and different ones is not clear at first sight. Finally, let's compare
the HPDIs for Reddit and Reddit debiased datasets. For Religion and Race
dataset there is a minor shift (in absolute values the highest change is
equal to approximately 0.1) of the mean coefficients towards zero.
However for Gender dataset there is no significant change. This is of
course a general look at the group coefficients. Let's now analyze the
individual words.

@@

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\hypertarget{ref-Manzini2019blackToCriminal}{}%
Manzini, T., Lim, Y. C., Tsvetkov, Y., \& Black, A. W. (2019).
\emph{Black is to criminal as caucasian is to police: Detecting and
removing multiclass bias in word embeddings}. Retrieved from
\url{http://arxiv.org/abs/1904.04047}

\leavevmode\hypertarget{ref-Venkat2018Curse}{}%
Venkat, N. (2018). \emph{The curse of dimensionality: Inside out}.
\url{https://doi.org/10.13140/RG.2.2.29631.36006}
